---
title: "QC_plots"
author: "Osman Sharifi"
date: '2022-07-26'
output: html_document
---
Derived from scCustomize 

```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)
```

# QC Plots
One of the first steps in all scRNA-seq analyses is performing a number of QC checks and plots so that data can be appropriately filtered.  scCustomize contains a number of functions that can be used to quickly and easily generate some of the most relevant QC plots.

For this tutorial, I will be utilizing HCA bone marrow cell data from the SeuratData package.
```{r init}
library(tidyverse)
library(patchwork)
library(Seurat)
library(scCustomize)
library(qs)

# Load Dataset
human_rettcort <- hcabm40k.SeuratData::hcabm40k
load(human_rettcort)
```


```{r include=FALSE}
accepted_names <- Add_Mito_Ribo_Seurat(list_species_names = TRUE)
```

## Add Mitochondrial and Ribosomal Gene Percentages
scCustomize contains easy wrapper function to automatically add both Mitochondrial and Ribosomal count percentages to meta.data slot.  If you are using mouse, human, or marmoset data all you need to do is specify the `species` parameter.
```{r}
# These defaults can be run just by providing accepted species name
human_rettcort <- Add_Mito_Ribo_Seurat(seurat_object = human_rettcort, species = "Human", overwrite = TRUE)
```
*NOTE: There is analogous function for LIGER objects (see: `Add_Mito_Ribo_LIGER`) but QC plotting functions currently only support Seurat objects.*

To view list of accepted values for default species names simply set `list_species_names = TRUE`.
```{r echo=FALSE}
accepted_names %>%
  kableExtra::kbl() %>%
  kableExtra::kable_styling(bootstrap_options = c("bordered", "condensed", "responsive", "striped")) 
```

However custom prefixes can be used for non-human/mouse/marmoset species with different annotations. Simply specify `species = other` and supply feature lists or regex patterns for your species of interest.  
*NOTE: If desired please submit issue on GitHub for additional default species.  Please include regex pattern or list of genes for both mitochondrial and ribosomal genes and I will add additional built-in defaults to the function.*
```{r eval=FALSE}
# Using gene name patterns
human_rettcort <- Add_Mito_Ribo_Seurat(seurat_object = human_rettcort, species = "other", mito_pattern = "regexp_pattern", ribo_pattern = "regexp_pattern")

# Using feature name lists
mito_gene_list <- c("gene1", "gene2", "etc")
ribo_gene_list <- c("gene1", "gene2", "etc")
human_rettcort <- Add_Mito_Ribo_Seurat(seurat_object = human_rettcort, species = "other", mito_features = mito_gene_list, ribo_features = ribo_gene_list)

# Using combination of gene lists and gene name patterns
human_rettcort <- Add_Mito_Ribo_Seurat(seurat_object = human_rettcort, species = "Human", mito_features = mito_gene_list, ribo_pattern = "regexp_pattern")
```

## Plotting QC Metrics
scCustomize has a number of quick QC plotting options for ease of use.  
*NOTE: Most scCustomize plotting functions contain `...` parameter to allow user to supply any of the parameters for the original Seurat function that is being used under the hood.*

## `VlnPlot`-Based QC Plots  
scCustomize contains 4 functions which wrap `Seurat::VlnPlot()`.  

* `QC_Plots_Genes()` Plots genes per cell/nucleus.
* `QC_Plots_UMIs()` Plots UMIs per cell/nucleus.  
* `QC_Plots_Mito()` Plots mito% (named "percent_mito") per cell/nucleus.
* `QC_Plots_Feature()` Plots "feature" per cell/nucleus.  Using parameter `feature` to allow plotting of any applicable named feature in object\@meta.data slot. 

scCustomize functions have the added benefit of:  

* Feature to plot set by default (except for `QC_Plots_Feature`).
* Added high/low cutoff parameters to allow for easy visualization of potential cutoff thresholds.


```{r eval=FALSE}
# All functions contain 
human_rettcort$percent_mito = human_rettcort$percent.mito
p1 <- QC_Plots_Genes(seurat_object = human_rettcort, low_cutoff = 800, high_cutoff = 5500)
p2 <- QC_Plots_UMIs(seurat_object = human_rettcort, low_cutoff = 1200, high_cutoff = 45000)
p3 <- QC_Plots_Mito(seurat_object = human_rettcort, high_cutoff = 20)
```

```{r echo=FALSE}
# All functions contain 
p1 <- QC_Plots_Genes(seurat_object = human_rettcort, low_cutoff = 800, high_cutoff = 5500, pt.size = 0.4)
p2 <- QC_Plots_UMIs(seurat_object = human_rettcort, low_cutoff = 1200, high_cutoff = 15000, pt.size = 0.4)
p3 <- QC_Plots_Mito(seurat_object = human_rettcort, high_cutoff = 5, pt.size = 0.4)
```

```{r, fig.height=7, fig.width=13}
wrap_plots(p1, p2, p3, ncol = 3)
ggplot2::ggsave("/Users/osman/Documents/GitHub/snRNA-seq-pipeline/scripts/07_human_cell_labeling/QC/vln_gene_umi_mito.pdf",
                device = NULL,
                height = 8.5,
                width = 20)

```

### Additional parameters  
In addition to being able to supply Seurat parameters with `...` these plots like many others in scCustomize contain other additional parameters to customize plot output without need for post-plot ggplot2 modifications  

* `plot_title`: Change plot title
* `x_axis_label`/`y_axis_label`: Change axis labels.
* `x_lab_rotate`: Should x-axis label be rotated 45 degrees?
* `y_axis_log`: Should y-axis in linear or log10 scale.


```{r, fig.height=5, fig.width=13, fig.align='center', fig.cap="*Setting `y_axis_log` can be very helpful for initial plots where outliers skew the visualization of the majority of the data without excluding data by setting y-axis limit.*"}
p1 <- QC_Plots_UMIs(seurat_object = human_rettcort, low_cutoff = 1200, high_cutoff = 45000, pt.size = 0.1)
p2 <- QC_Plots_UMIs(seurat_object = human_rettcort, low_cutoff = 1200, high_cutoff = 45000, pt.size = 0.1, y_axis_log = TRUE)

wrap_plots(p1, p2, ncol = 2)

```




## `FeatureScatter`-Based QC Plots  
scCustomize contains 3 functions which wrap `Seurat::FeatureScatter()` with added visualization of potential cutoff thresholds and some additional functionality:  

* `QC_Plot_UMIvsGene()` Plots genes vs UMIs per cell/nucleus
* `QC_Plot_GenevsFeature()` Plots Genes vs. "feature" per cell/nucleus.  Using parameter `feature1` to allow plotting of any applicable named feature in object\@meta.data slot.   
* `QC_Plot_UMIvsFeature()` Plots UMIs vs. "feature" per cell/nucleus.  Using parameter `feature1` to allow plotting of any applicable named feature in object\@meta.data slot. 

### New/Modified functionality  

* Better default color palettes
* `shuffle = TRUE` by default to prevent hiding of datasets
* Ability to set & visualize potential cutoff thresholds (similar to VlnPlot based QC Plots above)
* Report potential post filtering correlation in addition to whole dataset correlation when using `QC_Plot_UMIvsGene` (based on values provided to high and low cutoff parameters)

```{r eval=FALSE}
# All functions contain 
QC_Plot_UMIvsGene(seurat_object = human_rettcort, low_cutoff_gene = 800, high_cutoff_gene = 5500, low_cutoff_UMI = 500, high_cutoff_UMI = 15000)
ggplot2::ggsave("/Users/osman/Documents/GitHub/snRNA-seq-pipeline/scripts/07_human_cell_labeling/QC/.pdf",
                device = NULL,
                height = 8.5,
                width = 12)
QC_Plot_GenevsFeature(seurat_object = human_rettcort, feature1 = "percent_mito", low_cutoff_gene = 800, high_cutoff_gene = 5500, high_cutoff_feature = 5)
```

```{r echo=FALSE, fig.height=7, fig.width=13, fig.align='center'}
# All functions contain 
p1 <- QC_Plot_UMIvsGene(seurat_object = human_rettcort, low_cutoff_gene = 400, high_cutoff_gene = 5500, low_cutoff_UMI = 1200, high_cutoff_UMI = 15000, x_axis_label = "UMIs per Cell/Nucleus")
p2 <- QC_Plot_GenevsFeature(seurat_object = human_rettcort, feature1 = "percent_mito", low_cutoff_gene = 800, high_cutoff_gene = 5500, high_cutoff_feature = 5, low_cutoff_feature = 0)
p3 <- QC_Plot_GenevsFeature(seurat_object = human_rettcort, feature1 = "percent_ribo", low_cutoff_gene = 800, high_cutoff_gene = 5500, high_cutoff_feature = 8, low_cutoff_feature = 0)

wrap_plots(p1, p2, p3, ncol = 3)
ggplot2::ggsave("/Users/osman/Documents/GitHub/snRNA-seq-pipeline/scripts/07_human_cell_labeling/QC/QC_correlation.pdf",
                device = NULL,
                height = 8.5,
                width = 20)
```


### Color data by continuous meta data variable  
`QC_Plot_UMIvsGene` contains the ability to color points by continuous meta data variables.  
This can be used to plot % of mito reads in addition to UMI vs. Gene comparisons
```{r eval=FALSE}
QC_Plot_UMIvsGene(seurat_object = human_rettcort, meta_gradient_name = "percent_mito", low_cutoff_gene = 800, high_cutoff_gene = 5500, high_cutoff_UMI = 45000)
QC_Plot_UMIvsGene(seurat_object = human_rettcort, meta_gradient_name = "percent_mito", low_cutoff_gene = 800, high_cutoff_gene = 5500, high_cutoff_UMI = 15000, meta_gradient_low_cutoff = 5)
```

```{r echo=FALSE, fig.height=7, fig.width=13, fig.cap="*`QC_Plot_UMIvsGene()` when using `meta_gradient_name` outputs plot colored by meta data variable (left) to view only points above potential cutoff `meta_gradient_low_cutoff` can be specified to alter the plotting (right).*", fig.align='center'}
p1 <- QC_Plot_UMIvsGene(seurat_object = human_rettcort, meta_gradient_name = "percent_mito", low_cutoff_gene = 800, high_cutoff_gene = 5500, high_cutoff_UMI = 45000, meta_gradient_low_cutoff = 5)
p2 <- QC_Plot_UMIvsGene(seurat_object = human_rettcort, meta_gradient_name = "percent_mito", low_cutoff_gene = 800, high_cutoff_gene = 5500, high_cutoff_UMI = 45000, meta_gradient_low_cutoff = 5)

wrap_plots(p1, p2, ncol = 2)
```

### Combination Plots
If you are interested in viewing `QC_Plot_UMIvsGene` both by discrete grouping variable and by continuous variable without writing function twice you can use `combination = TRUE` and plot output will contain both plots.
```{r, fig.height=7, fig.width=13, fig.cap="*`QC_Plot_UMIvsGene()` when using `combination = TRUE` will output both the Gene x UMI by active identity and with meta data gradient coloring.*", fig.align='center'}
QC_Plot_UMIvsGene(seurat_object = human_rettcort, meta_gradient_name = "percent_ribo", low_cutoff_gene = 800, high_cutoff_gene = 5500, high_cutoff_UMI = 15000, meta_gradient_low_cutoff = 10, combination = TRUE)
```

## Analyze Median QC Values per Sample/Library
scCustomize also contains a few helpful functions for returning and plotting the median values for these metrics on per sample/library basis.

### Calculate Median Values & Return data.frame
scCustomize contains function `Median_Stats` to quickly calculate the medians for basic QC stats (Genes/, UMIs/, %Mito/Cell, etc) and return a data.frame.
```{r}
median_stats <- Median_Stats(seurat_object = human_rettcort, group_by_var = "Samples")
```


```{r echo=FALSE}
median_stats %>%
  kableExtra::kbl() %>%
  kableExtra::kable_styling(bootstrap_options = c("bordered", "condensed", "responsive", "striped")) 
```


The `Median_Stats` function has some column names stored by default but will also calculate medians for additional meta.data columns using the optional `median_var` parameter
```{r eval=FALSE}
median_stats <- Median_Stats(seurat_object = human_rettcort, group_by_var = "orig.ident", median_var = "meta_data_column_name")
```


### Plotting Median Values
scCustomize also contains a few functions to plot some of these median value calculations, which can be used on their own without need to return data.frame first.

* `Plot_Median_Genes()`  
* `Plot_Median_UMIs()`  
* `Plot_Median_Mito()`  
* `Plot_Median_Other()`  
  - *Used to plot any other numeric variable present in object meta.data slot.*
```{r eval=FALSE}
Plot_Median_Genes(seurat_object = human_rettcort, group_by = "predicted.class")
Plot_Median_UMIs(seurat_object = human_rettcort, group_by = "group")
Plot_Median_Mito(seurat_object = human_rettcort, group_by = "group")
Plot_Median_Other(seurat_object = human_rettcort, median_var = "percent_ribo")
```


```{r echo=FALSE, fig.align='center', fig.height=10, fig.width=13}
p1 <- Plot_Median_Genes(seurat_object = human_rettcort, group_by = "group")
p2 <- Plot_Median_UMIs(seurat_object = human_rettcort, group_by = "group")
p3 <- Plot_Median_Mito(seurat_object = human_rettcort, group_by = "group")
p4 <- Plot_Median_Other(seurat_object = human_rettcort, median_var = "percent_ribo", group_by = "group")

wrap_plots(p1, p2, p3, p4, ncol = 2)
```

### Plot Number of Cells/Nuclei per Sample
scCustomize also contains plotting function to plot the number of cells or nuclei per sample.

Since the HCA Bone Marrow dataset has exactly the same number of cells per sample we will use the microglia object from the Analysis Plots vignette.

```{r include=FALSE}
marsh_mouse_micro <- qread(file = "assets/marsh_2020_micro.qs")
```


```{r echo=FALSE, fig.align='center', fig.height=8.5, fig.width=12}
scCustomize::Plot_Cells_per_Sample(seurat_object = human_rettcort, group_by = "Samples")
ggplot2::ggsave("/Users/osman/Documents/GitHub/snRNA-seq-pipeline/scripts/07_human_cell_labeling/QC/cellpersample.pdf",
                device = NULL,
                height = 8.5,
                width = 12)
levels(human_rettcort$Samples) <- c("1136_CTRL", "1711_CTRL", "1846_CTRL", "5125_CTRL", "5161_CTRL", "662_CTRL", "4591_CTRL", "812_CTRL","1420_RTT", "1815_RTT", "3381_RTT", "4321_RTT", "4687_RTT", "4852_RTT", "5020_RTT")
```
```{r echo=FALSE, fig.align='center', fig.height=8.5, fig.width=12}
#select cells with percent.mito < 25
selected <- WhichCells(human_rettcort, expression = percent_mito < 5)
human_rett_cort_filt <- subset(human_rettcort, cells = selected)
selected <- WhichCells(human_rettcort, expression = nFeature_RNA > 800)
human_rett_cort_filt <- subset(human_rett_cort_filt, cells = selected)
selected <- WhichCells(human_rett_cort_filt, expression = nFeature_RNA < 5500)
human_rett_cort_filt <- subset(human_rett_cort_filt, cells = selected)
selected <- WhichCells(human_rett_cort_filt, expression = nCount_RNA > 1200)
human_rett_cort_filt <- subset(human_rett_cort_filt, cells = selected)
selected <- WhichCells(human_rett_cort_filt, expression = nCount_RNA < 15000)
human_rett_cort_filt <- subset(human_rett_cort_filt, cells = selected)
selected <- WhichCells(human_rett_cort_filt, expression = percent_ribo < 8)
human_rett_cort_filt <- subset(human_rett_cort_filt, cells = selected)
length(selected)


p1 <- QC_Plot_UMIvsGene(seurat_object = human_rett_cort_filt, low_cutoff_gene = 400, high_cutoff_gene = 5500, low_cutoff_UMI = 1200, high_cutoff_UMI = 15000, x_axis_label = "UMIs per Cell/Nucleus")
p2 <- QC_Plot_GenevsFeature(seurat_object = human_rett_cort_filt, feature1 = "percent_mito", low_cutoff_gene = 800, high_cutoff_gene = 5500, high_cutoff_feature = 5, low_cutoff_feature = 0)
p3 <- QC_Plot_GenevsFeature(seurat_object = human_rett_cort_filt, feature1 = "percent_ribo", low_cutoff_gene = 800, high_cutoff_gene = 5500, high_cutoff_feature = 8, low_cutoff_feature = 0)
wrap_plots(p1, p2, p3, ncol = 3)
ggplot2::ggsave("/Users/osman/Documents/GitHub/snRNA-seq-pipeline/scripts/07_human_cell_labeling/QC/QC_correlation_filt.pdf",
                device = NULL,
                height = 8.5,
                width = 20)

post <- scCustomize::Plot_Cells_per_Sample(seurat_object = human_rett_cort_filt, group_by = "Samples")
ggplot2::ggsave("/Users/osman/Documents/GitHub/snRNA-seq-pipeline/scripts/07_human_cell_labeling/QC/cellpersample_filt.pdf",
                device = NULL,
                height = 8.5,
                width = 12)
```
